### 2008-03-20 trunk/Papers/dl-gre-08/bisim.tex
Written the algorithm for ALC and changed the one for EL.
Main changes in Section 2 and in the algorithms. Intro in section 2 completed.  Still working in the end of Section 2.  Section 3 only changed by addition of text that was in Section 2 before. I still have to work here.
In addition, added some references to the bib.

In computational linguistics, the generating referring expression problem (the GRE
problem) can be intuitively presented as follows: given some information about
a group of individuals (e.g., grounded facts about individuals in an ongoing dialogue),
and a grammar, found a grammatically correct expression (usually a noun phrase) that uniquely identify a given individual.

By taking a logic perspective, we can recast the GRE problem as an inference problem.
We can think that the information about the group of individuals is provided in the form
of a model $\gM$ of a given logical language $\gL$.  Now, the inference problem
associated with the GRE problem will be to find a formula in $\gL$ such that it
exacly identify a given individual $i$.  Formally, if $|\cdot|^\gM$ is the interpretation
function for $\gL$ which for each formula in $\gL$ and for each model $\gM$ returns
the set of elements in $\gM$ for which the formula $\varphi$ holds, then we can define
\medskip

\noindent
{\small
\begin{tabular}{l} \hline
\textsc{$\gL$-GRE Problem}\\ \hline
\ \ \ Input: A model $\gM$, an individual $i$ in the domain of $\gM$\\
\ \ \ Output: A formula $\varphi \in \gL$ such that $|\varphi|^\gM = \{i\}$\\
\hspace*{1.4cm} (if such a formula exists).\\ \hline
\end{tabular}}
\medskip

As it can be seen from the definition, the problem depends on which particular
logical language $\gL$ we are interested in, and (as the original GRE problem)
it might terminate with failure.  Depending on the model $\gM$, the chosen individual $i$
and the expressive power of $\gL$ it can be the case that there is no formula of
$\gL$ that uniquely identifies $i$.

Consider the following example.  We have three individuals related as indicated in
Figure~\ref{fig-car1}.
\begin{figure}
\begin{center}
\includegraphics[scale=.8]{pic.pdf}
\end{center}
\caption{Distinguishable/indistinguishable points.}\label{fig-car1}
\end{figure}
%
The individuals $i$ and $i'$ can easily be distinguished in first-order logic:
the formula $R(x,x)$ is true of $i$ and false in $i'$. The corresponding
referring expression for $i$ would be ``an individual relating to itself''.
But if we can only use propositional and relational information (equivalently, our grammar does not contain the word ``itself'') then $i$ and $i'$ are indistinguishable. Both
are ``individuals which are related to individuals which are related to individuals\ldots''

In what follows we will introduce two logical languages (called \el and \alc, see~\cite{baad:desc03}) which can naturally be mapped to standard grammatically correct expressions.  \el allows
for conjunctions of atomic and existentially quantified relational expressions (e.g.,
``a red book on a table'') while \alc is the extension of \el with full negation
(e.g., ``a book which is not red and is not on the floor'').

@@

We can now exactly characterize the problem of when a formula distinguishing two
individuals in a given model exists. We will do so using the notion of \emph{bisimulation}.

@@

\begin{theorem}[\cite{blac:moda01}]\label{bisim}
Given a model $\gM$, if  $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
then there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
$i' \not \in |\varphi|^\gM$.
\end{theorem}

>>>>

One of the main ideas we want to investigate in this paper is
how we can use \emph{formulas} from description logics (see~\cite{baad:desc03}) to represent
referring expressions (RE). And how we can use the notion of \emph{bisimulation}
(see~\cite{blac:moda01}) to obtain very efficient algorithms for computing referring expressions.

In what follows we will introduce in detail two description logics called \el and \alc.  Intuitively, \el allows
for conjunctions of atomic and existentially quantified relational expressions (e.g.,
we can write formulas in \el that could be realized as ``a red book on a table''). \alc is the extension of \el with full negation
(e.g., we can write formulas in \alc which could be realized as ``a book which is not red and is not on the floor'').  Computing bisimulations in \alc is very efficient, but \alc can contain expressions which are difficult to realize (because of the presence of negation).  On the other hand,
\el expressions are easily realizable, but the bisimulation algorithm is more complex.

But let start by formally defining the description languages we are going to use, and the notion of bisimulation.

@@

The previous definition formally introduces the syntax and semantics of the
logics \alc and \el.  In the paper we will sometime refer to some other description logics (mainly in Section~\ref{xxx}).  For completeness we mention them briefly here,
and we refer the reader to~\cite{baad:desc03} for further details:

\begin{center}
\begin{tabular}{lp{6cm}}
$\mathcal{CL}$ & is obtained from $\el$ by dropping existential quantification\\
$\mathcal{PL}$ & is obtained form $\alc$ by dropping existential quantification\\
$\mathcal{ELU}_{(\neg)}$ & is obtained from $\el$ by adding disjunction and atomic negation
\end{tabular}
\end{center}

\noindent
It should be clear from the definitions above that $\mathcal{CL}$ is conjunctive logic, while $\mathcal{PL}$ is propositional logic.

We now introduce the notion of \emph{bisimulation}.  We will discuss in detail
the case for the $\alc$ language.  Further details can be found
in~\cite{blac:moda01,kurt:expr99}.

@@

\begin{theorem}[]\label{bisim}
Given a model $\gM$, if  $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
then there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
$i' \not \in |\varphi|^\gM$.
\end{theorem}

### 2008-03-20 trunk/Papers/dl-gre-08/gre.tex
Main changes in Section 2 and in the algorithms. Intro in section 2 completed.  Still working in the end of Section 2.  Section 3 only changed by addition of text that was in Section 2 before. I still have to work here.
In addition, added some references to the bib.

@@

\begin{algorithm}[t]
\dontprintsemicolon
\caption{\el\ bisimulation classes}
\label{algo:bisim-el}
$\C \leftarrow \{\top\}$ \;
\For{$p \in \prop$}{
  add$_\el$(p, \C) \;
}
\While{not all concepts in $\C$ singleton}{
  \For{$C \in \C$, $R \in \rel$}{
    add$_\el$($\exists R.C$, $\C$) \;
  }
  \If{made no changes to $\C$}{
    exit\;
  }
}
\end{algorithm}

\begin{algorithm}[t]
\dontprintsemicolon
\caption{add$_\el$($C$, $\C$)}
\label{algo:bisim-add-el}
\For{$C' \in \C$ non-singleton}{
  \If{$C' \sqcap C$ is not subsumed in $\C$}{
    add $C' \sqcap C$ to $\C$ \;
    remove subsumed concepts from $\C$\;
  }
}
\end{algorithm}

>>>>

In computational linguistics, the generating referring expression problem (the GRE
problem) can be intuitively presented as follows: given some information about
a group of individuals (e.g., grounded facts about individuals in an ongoing dialogue),
and a grammar, found a grammatically correct expression (usually a noun phrase) that uniquely identify a given individual.

By taking a logic perspective, we can recast the GRE problem as an inference problem.
We can think that the information about the group of individuals is provided in the form
of a model $\gM$ of a given logical language $\gL$.  Now, the inference problem
associated with the GRE problem will be to find a formula in $\gL$ such that it
exaclty identifies a given individual $i$.  Formally, if $|\cdot|^\gM$ is the interpretation
function for $\gL$ which for each formula in $\gL$ and for each model $\gM$ returns
the set of elements in $\gM$ for which the formula $\varphi$ holds, then we can define:
\medskip

\noindent
{\small
\begin{center}
\begin{tabular}{ll} \hline
\multicolumn{2}{l}{
\textsc{$\gL$-GRE Problem}}\\ \hline
\ \ Input: & A model $\gM$ and an individual $i$ in the\\
& \hspace*{0.5cm} domain of $\gM$.\\
\ \ Output: & A formula $\varphi \in \gL$ such that $|\varphi|^\gM = \{i\}$\\
& \hspace*{0.5cm} (if such a formula exists).\\ \hline
\end{tabular}
\end{center}}
\medskip

As it can be seen from the definition, the problem depends on which particular
logical language $\gL$ we are interested in, and (as the original GRE problem)
it might terminate with failure.  Depending on \emph{a)} the model $\gM$,
\emph{b)} the chosen individual $i$
and \emph{c} the expressive power of $\gL$, it can be the case that there is no formula of
$\gL$ that uniquely identifies $i$ in $\gM$.

Consider the following example.  We have three individuals related as indicated in
Figure~\ref{fig-car1}.
\begin{figure}
\begin{center}
\includegraphics[scale=.8]{pic.pdf}
\end{center}
\caption{Distinguishable/indistinguishable points.}\label{fig-car1}
\end{figure}
%
The individuals $i$ and $i'$ can easily be distinguished in first-order logic:
the formula $R(x,x)$ is true of $i$ and false in $i'$. The corresponding
referring expression for $i$ would be ``an individual relating to itself''.
But if we can only use propositional and relational information (equivalently, our grammar does not contain the word ``itself'') then $i$ and $i'$ are indistinguishable. Both
are ``individuals which are related to individuals which are related to individuals\ldots''

@@

% \begin{algorithm}[t]
% \dontprintsemicolon
% \caption{\el\ bisimulation classes}
% \label{algo:bisim-el}
% \KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
% \KwOut{A set \RE of pairs $(\varphi,|\varphi|^\gM)$ such that
% $\{S \mid (\varphi,S) \in \RE\}$ is the set of $\el$-bisimulation
% classes of $\gM$.}
% \RE $\leftarrow \{(\top,\Delta^\gM)\}$ \;
% \For{$p \in \prop$}{
%   add$_\el$(p, \RE) \;
% }
% \While{for some $(\varphi,S) \in \RE$, $|S| > 1$}{
%   \For{$(\varphi,S) \in$ \RE, $R \in \rel$}{
%     add$_\el$($\exists R.\varphi$, \RE) \;
%   }
%   \If{made no changes to $\RE$}{
%     exit\;
%   }
% }
% \end{algorithm}

\begin{algorithm}[t]
\dontprintsemicolon
\caption{}
\label{algo:bisim-alc}
\KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
\KwOut{A set \RE of formulas  such that
$\{|\varphi|^\gM \in \RE\}$ is the set of $\mathcal{L}$-bisimulation
classes of $\gM$.}

$\RE \leftarrow \{\top\}$

\For{$p \in \prop$}{
      add$_\mathcal{L}(p,\RE)$
   }

\While{for some $\varphi \in \RE, |\varphi|^\gM>1$}{
   \For{$\varphi \in \RE, R \in \rel$}{
         add$_\mathcal{L}(\exists R.\varphi,\RE)$
   }
   \If{made no changes to \RE}{
      exit
      }
}

\end{algorithm}

\begin{algorithm}
\caption{add$_\alc(\varphi,\RE)$}
\For{$\psi \in \RE$ with $|\psi|^\gM > 1$}{
   \If{$|\psi \sqcap \varphi|^\gM \not = \emptyset$ and
       $|\psi \sqcap \neg \varphi|^\gM \not = \emptyset$}{
         add $\psi \sqcap \varphi$ and
               $\psi \sqcap \neg \varphi$ to \RE \;
         remove $\psi$ from \RE \;
      }
   }
\end{algorithm}

\begin{algorithm}[t]
\dontprintsemicolon
\caption{add$_\el$($\varphi$, $\RE$)}
\label{algo:bisim-add-el}
\For{$\psi \in \RE$ with $|\psi|^\gM > 1$}{
  \If{$\psi \sqcap \varphi$ is not subsumed in $\RE$ {\bf and}
    $|\psi \sqcap \varphi|^\gM \neq \emptyset$ {\bf and}
    $|\psi \sqcap \varphi|^\gM \neq |\psi|^\gM$}{
    add $\psi \sqcap \varphi$ to $\RE$ \;
    remove subsumed concepts from $\RE$\;
  }
}
\end{algorithm}
